Model Finding for Recursive Functions in SMT

### 2015-04-28 tex/main.tex
Minor fix
Work on results section, minor changes.

\title{Model Finding for Finite Function Traces in SMT}

@@

\subsection{Preliminaries}

A \emph{signature} $\Sigma$ consists of 
a set $\ssorts{\Sigma}$ of sort symbols and
a set $\sfuns{\Sigma}$ of function symbols.
We assume that signatures always include a Boolean sort $\Bool$ and constants 
$\ltrue$ and $\lfalse$ of that sort (respectively, for true and false), 
an (infix) equality predicate $\teq$ of type $\sigma \times \sigma$ for each $\sigma \in \ssorts{\Sigma}$,
and standard boolean connectives ($\neg$, $\wedge$, $\vee$, $\lite$, etc.).
%and a function $\lite$ of type $\Bool \times \sigma \times \sigma$ for each $\sigma \in \ssorts{\Sigma}$.
We will write $t\typ{\tau}$ to denote a term $t$ of type~$\tau$.

@@

\ajr{theories, models, etc.}

\begin{definition}
A formula $\varphi$ \emph{well-defined with respect to functions} $S$ if it is of the form:
\begin{equation*}
(\forallf{f_1} \vec x_1. \varphi_1) \wedge \ldots \wedge (\forallf{f_n} \vec x_n. \varphi_n) \wedge \psi
\end{equation*}
where $f_1, \ldots, f_n$ are distinct, 
$S = \{ f_1, \ldots, f_n \}$,
and $\psi$ contains no function definitions.
\end{definition}

@@

\subsection{Encoding Well-Defined Assumption for Recursive Functions}
$\Leftrightarrow$, 
In the remainder of the section, we fix a signature $\Sigma$.

@@

\begin{figure}[t]
\begin{enumerate}
%\begin{framed}
\item[\ ] 
$\conv( t\typ{\tau}, p )$ : 
 \begin{itemize}
   \item[] $\mathsf{if}$ $\tau \equiv \Bool$ and $t \equiv t_1 \boolop \ldots \boolop t_n$ :
    \begin{itemize}
      \item[] $\mathsf{let}$ $( r_i, D_i ) = \conv( t_i, \pol( \boolop, i, p ) )$ for $i = 1, \ldots, n$
      \item[] $\mathsf{if}$ $p = \ptrue$
      \begin{itemize}
        \item[] $\mathsf{return}$ $( ( r_1 \boolop \ldots \boolop r_n ) \wedge D_1 \wedge \ldots \wedge D_n, \emptyset )$
      \end{itemize}
      \item[] $\mathsf{else}$ $\mathsf{if}$ $p = \pfalse$
      \begin{itemize}
        \item[] $\mathsf{return}$ $( ( r_1 \boolop \ldots \boolop r_n ) \vee \neg D_1 \vee \ldots \vee \neg D_n, \emptyset )$
      \end{itemize}
      \item[] $\mathsf{else}$
      \begin{itemize}
        \item[] $\mathsf{return}$ $( r_1 \boolop \ldots \boolop r_n, D_1 \cup \ldots \cup D_n )$
      \end{itemize}
    \end{itemize}
  \item[] $\mathsf{else}$ $\mathsf{if}$ $t \equiv \forallf{f} \vec x. t_1$
    \begin{itemize}
      \item[] $\mathsf{let} ( r_1, D_1 ) = \conv( t_1, p )$
      \item[] $\mathsf{if}$ $f \equiv \fnull$
      \begin{itemize}
        \item[] $\mathsf{return}$ $( \forall \vec x. r_1, \forall \vec x. D_1 )$
      \end{itemize}
      \item[] else
      \begin{itemize}
        \item[] $\mathsf{assert}( D_1 = \emptyset )$;
        \item[] $\mathsf{return}$ $( \forall y\typ{\fargsort{f}}. ( r_1 [ \vecfarg{f}( y ) / \vec x ] ), \emptyset )$
      \end{itemize}
    \end{itemize}
   \item[] $\mathsf{else}$
   \begin{itemize}
     \item[] $\mathsf{return}$ $( t, \{ \exists z\typ{\fargsort{f}}. ( \vecfarg{f}( z ) \teq \vec t ) \mid f( \vec t ) \in \functerms^\Sigma( t ) \} )$
   \end{itemize}
 \end{itemize}
\end{enumerate}
\vspace{-2ex}
\caption{A translation procedure $\conv$ for a $\Sigma$-formula that is well-defined with respect to function definitions.
In the procedure, $\pol( \boolop, i, p )$ returns the polarity of the $i^{th}$ child of a $\boolop$-application having polarity $p$,
where $\boolop$ is an (interpreted) predicate.
}
\label{fig:encoding}
\end{figure}

The intuition behind this translation is use an uninterpreted sort $\fargsort{f}$ to abstract the set of tuples that each recursive function $f$ is invoked on,
and restrict the quantification of the function definition for $f$ to a variable of sort $\fargsort{f}$.
In more detail, we construct an extended signature of $\Sigma$, which we will write as $\extendsig{\Sigma}$, as follows.
For each $f \in \sfundefs{\Sigma}$ of type $\tau_1 \times \ldots \tau_n \rightarrow \tau$, 
the signature $\extendsig{\Sigma}$ includes:
\begin{enumerate}
\item[-] an uninterpreted sort $\fargsort{f}$, and
\item[-] a vector of $n$ uninterpreted functions $\farg{f}^1$, $\ldots$, $\farg{f}^n$ of type $(\fargsort{f} \rightarrow \tau_1)$, $\ldots$, $(\fargsort{f} \rightarrow \tau_n)$.
\end{enumerate}
%The interpretation of uninterpreted sort $\fargsort{f}$ will denote the elements (tuples) on which the function $f$ is applied.
%The role of the uninterpreted functions $\farg{f}^1$, $\ldots$, $\farg{f}^n$ will be discussed more in the following.
We then run the procedure $\conv$ in Figure~\ref{fig:encoding} on $\varphi$.
The procedure $\conv$ takes two arguments: the term $t$ to translate, and a \emph{polarity} $p$ (either $\ptrue$, $\pfalse$, or $\pnone$).
It returns a pair of the form $( r, D )$, where $r$ is a term of type $\tau$, and $D$ is a set of formulas, which we will call its \emph{domain constaints}.
The role of $D$ is to ensure that the (restricted) function definition for $f$ includes certain tuples in its domain, which we explain more in the following.

In the case that $t$ is an application of (interpreted) predicate $\boolop$ (e.g. a boolean connective, or equality),
we recursively call $\conv$ on its subchildren $t_i$ and polarity $\pol( \boolop, i, p )$, where $\pol$ is defined as\footnote{The negation $\neg p$ of polarity $p$ is 
$\pfalse$ if $p$ is $\ptrue$,
$\ptrue$ if $p$ is $\pfalse$, and
$\pnone$ if $p$ is $\pnone$.}:
\begin{equation*}
\pol( \boolop, i, p ) = \begin{cases}
                         \pnone & \boolop \equiv \teq \text{ or } \boolop \equiv \lite \wedge i=0 \\
                         \neg p & \boolop \equiv \neg \\
                         p & \text{otherwise}
                         \end{cases}
\end{equation*}
If $t$ has positive polarity,
we construct a conjunction of the result of the recursive calls with domain constraints $D_1, \ldots, D_n$.
Similarly, if $t$ has negative polarity,
we construct a disjunction of the result of the recursive calls with the negation of domain constraints $D_1, \ldots, D_n$.
It $t$ has no polarity,
then we return the union of the domain constraints $D_1 \cup \ldots \cup D_n$.

@@

\begin{example}
Consider the (combined) signature $\Sigma = \Sigmalia \cup \Sigma_u$,
where $\ssorts{\Sigma_u} = \{ \sortint \}$, and 
$\sfuns{\Sigma_u} = \{ s\typ{ \sortint \rightarrow \sortint }, c\typ{ \sortint } \}$,
and the $\Sigma$-formula $\varphi$:
\begin{equation*}
\forall_{s} x\typ{\sortint}. ite( x \leq 0, s(x) \teq 0, s( x ) \teq x + s( x-1 ) ) \wedge s( c ) > 100
\end{equation*}
Here, $s$ is a function that returns the sum of positive numbers between $0$ and its argument $x$.
The formula $\varphi$ is well-defined with respect to functions $\sfundefs{\Sigma_u}$,
and states that the sum of all positive numbers between $0$ and $c$ is greater than $100$.
The signature $\extendsig{\Sigma}$ includes the sort $\fargsort{s}$,
and the uninterpreted function $\farg{s}\typ{ \fargsort{s} \rightarrow \sortint }$.
The result of $\conv( \varphi, \ptrue )$ is the pair $( \psi, \emptyset )$, where after simplification, $\psi$ is the $\extendsig{\Sigma}$-formula:
\begin{equation*}
\begin{split}
\forall y\typ{\fargsort{s}}. ite( & \farg{s}( y ) \leq 0,  \\
 & s(\farg{s}( y )) \teq 0, \\
 & s(\farg{s}( y )) \teq \farg{s}( y ) + s( \farg{s}( y )-1 ) ) \\
 & \wedge \exists z\typ{\fargsort{s}}. \farg{s}( z ) \teq \farg{s}( y )-1 )
\end{split}  
\wedge ( s( a ) > 100 \wedge \exists z\typ{\fargsort{s}}. \farg{s}( z ) \teq a )
\end{equation*}
Note that the universal quantification in this example (the left disjunct) is $\ldots$
$\square$
\end{example}

We claim that if $\varphi$ is well-defined with respect to functions $\sfundefs{\Sigma}$,
function definitions in $\sfundefs{\Sigma}$ are logically consistent \ajr{formalize},
and $\conv( \varphi, \ptrue )$ returns $( \psi, \emptyset )$,
then $\psi$ is equisatisfiable to $\varphi$.

\begin{lemma}
Let $\varphi$ be a $\Sigma$-formula that is well-defined with respect to functions $\sfundefs{\Sigma}$,
and let $( \psi, D )$ be the result of $\conv( \varphi, \ptrue )$.
Then,
(i) $D = \emptyset$, and
(ii) for each model of $\varphi$, $\ldots$. \ajr{todo}
\end{lemma}

\section{Experiments}

@@

{
\bibliographystyle{abbrv}
\bibliography{bib}
}

>>>>

\title{Model Finding for Recursive Functions in SMT}

@@

\subsection{Preliminaries}
\label{sec:prelim}

A \emph{signature} $\Sigma$ consists of 
a set $\ssorts{\Sigma}$ of type symbols and
a set $\sfuns{\Sigma}$ of function symbols.
We assume that signatures always include a Boolean type $\Bool$ and constants 
$\ltrue$ and $\lfalse$ of that type (respectively, for true and false), 
an (infix) equality predicate $\teq$ of type $\sigma \times \sigma$ for each $\sigma \in \ssorts{\Sigma}$,
and standard boolean connectives ($\neg$, $\wedge$, $\vee$, $\lite$, etc.).
%and a function $\lite$ of type $\Bool \times \sigma \times \sigma$ for each $\sigma \in \ssorts{\Sigma}$.
We will write $t\typ{\tau}$ to denote a term $t$ of type~$\tau$.

@@

\ajr{theories, models, equisatisfiable, etc.}

\begin{definition}
A formula $\varphi$ \emph{well-defined with respect to functions} $S$ if it is of the form:
\begin{eqnarray} \label{eq:wdf}
(\forallf{f_1} \vec x_1. \varphi_1) \wedge \ldots \wedge (\forallf{f_n} \vec x_n. \varphi_n) \wedge \psi
\end{eqnarray}
where $f_1, \ldots, f_n$ are distinct, 
$S = \{ f_1, \ldots, f_n \}$,
and $\psi$ contains no function definitions.
\end{definition}

@@

\subsection{Encoding Well-Defined Assumption for Recursive Functions}
\label{sec:encoding}

In the remainder of the section, we fix a signature $\Sigma$.

@@

\begin{figure}[t]
\begin{enumerate}
%\begin{framed}
\item[\ ] 
$\conv( \varphi )$ : 
 \begin{itemize}
   \item[] $\mathsf{let}$ $( \psi, D ) = \conv( \varphi, \ptrue )$;
   %\item[] $\mathsf{assert}( D = \emptyset )$;
   \item[] $\mathsf{return}$ $\psi$
  \end{itemize}
\end{enumerate}
\begin{enumerate}
%\begin{framed}
\item[\ ] 
$\conv_0( t\typ{\tau}, p )$ : 
 \begin{itemize}
   \item[] $\mathsf{if}$ $\tau \equiv \Bool$ and $t \equiv t_1 \boolop \ldots \boolop t_n$ :
    \begin{itemize}
      \item[] $\mathsf{let}$ $( r_i, D_i ) = \conv_0( t_i, \pol( \boolop, i, p ) )$ for $i = 1, \ldots, n$;
      \item[] $\mathsf{if}$ $p = \ptrue$
      \begin{itemize}
        \item[] $\mathsf{return}$ $( ( r_1 \boolop \ldots \boolop r_n ) \wedge D_1 \wedge \ldots \wedge D_n, \emptyset )$
      \end{itemize}
      \item[] $\mathsf{else}$ $\mathsf{if}$ $p = \pfalse$
      \begin{itemize}
        \item[] $\mathsf{return}$ $( ( r_1 \boolop \ldots \boolop r_n ) \vee \neg D_1 \vee \ldots \vee \neg D_n, \emptyset )$
      \end{itemize}
      \item[] $\mathsf{else}$
      \begin{itemize}
        \item[] $\mathsf{return}$ $( r_1 \boolop \ldots \boolop r_n, D_1 \cup \ldots \cup D_n )$
      \end{itemize}
    \end{itemize}
  \item[] $\mathsf{else}$ $\mathsf{if}$ $t \equiv \forallf{f} \vec x. t_1$
    \begin{itemize}
      \item[] $\mathsf{let} ( r_1, D_1 ) = \conv_0( t_1, p )$;
      \item[] $\mathsf{if}$ $f \equiv \fnull$
      \begin{itemize}
        \item[] $\mathsf{return}$ $( \forall \vec x. r_1, \forall \vec x. D_1 )$
      \end{itemize}
      \item[] else
      \begin{itemize}
        %\item[] $\mathsf{assert}( D_1 = \emptyset )$;
        \item[] $\mathsf{return}$ $( \forall y\typ{\fargsort{f}}. ( r_1 [ \vecfarg{f}( y ) / \vec x ] ), \emptyset )$
      \end{itemize}
    \end{itemize}
   \item[] $\mathsf{else}$
   \begin{itemize}
     \item[] $\mathsf{return}$ $( t, \{ \exists z\typ{\fargsort{f}}. ( \vecfarg{f}( z ) \teq \vec t ) \mid f( \vec t ) \in \functerms^\Sigma( t ) \} )$
   \end{itemize}
 \end{itemize}
\end{enumerate}
\vspace{-2ex}
\caption{A translation procedure $\conv$ for a $\Sigma$-formula that is well-defined with respect to function definitions.
In the subprocedure $\conv_0$, the procedure $\pol( \boolop, i, p )$ returns the polarity of the $i^{th}$ child of a $\boolop$-application having polarity $p$,
where $\boolop$ is an (interpreted) predicate.
}
\label{fig:encoding}
\end{figure}

The intuition behind this translation is use an uninterpreted type $\fargsort{f}$ to abstract the set of tuples that each recursive function $f$ is invoked on,
and restrict the quantification of the function definition for $f$ to a variable of type $\fargsort{f}$.
In more detail, we construct an extended signature of $\Sigma$, which we will write as $\extendsig{\Sigma}$, as follows.
For each $f \in \sfundefs{\Sigma}$ of type $\tau_1 \times \ldots \tau_n \rightarrow \tau$, 
the signature $\extendsig{\Sigma}$ includes:
\begin{enumerate}
\item[-] an uninterpreted type $\fargsort{f}$, and
\item[-] a vector of $n$ uninterpreted functions $\farg{f}^1$, $\ldots$, $\farg{f}^n$ of type $(\fargsort{f} \rightarrow \tau_1)$, $\ldots$, $(\fargsort{f} \rightarrow \tau_n)$.
\end{enumerate}
%The interpretation of uninterpreted sort $\fargsort{f}$ will denote the elements (tuples) on which the function $f$ is applied.
%The role of the uninterpreted functions $\farg{f}^1$, $\ldots$, $\farg{f}^n$ will be discussed more in the following.
We then run the procedure $\conv$ in Figure~\ref{fig:encoding} on $\varphi$.
This calls the subprocedure $\conv_0$ on $\varphi$ and the \emph{polarity} $\ptrue$.
The procedure $\conv_0$ takes two arguments: the term $t$ to translate, and a polarity $p$ (either $\ptrue$, $\pfalse$, or $\pnone$).
It returns a pair of the form $( r, D )$, where $r$ is a term of type $\tau$, and $D$ is a set of formulas, which we will call its \emph{domain constaints}.
The role of $D$ is to ensure that the (restricted) function definition for $f$ includes certain tuples in its domain, which we explain more in the following.

In the case that $t$ is an application of (interpreted) predicate $\boolop$ (e.g. a boolean connective, or equality),
we recursively call $\conv$ on its subchildren $t_i$ and polarity $\pol( \boolop, i, p )$, where $\pol$ is defined as\footnote{The negation $\neg p$ of polarity $p$ is 
$\pfalse$ if $p$ is $\ptrue$,
$\ptrue$ if $p$ is $\pfalse$, and
$\pnone$ if $p$ is $\pnone$.}:
\begin{equation*}
\pol( \boolop, i, p ) = \begin{cases}
                         \pnone & \boolop \equiv \teq \text{ or } ( \boolop \equiv \lite \text{ and } i=0 ) \\
                         \neg p & \boolop \equiv \neg \\
                         p & \text{otherwise}
                         \end{cases}
\end{equation*}
If $t$ has positive polarity,
we construct a conjunction of the result of the recursive calls with domain constraints $D_1, \ldots, D_n$.
Similarly, if $t$ has negative polarity,
we construct a disjunction of the result of the recursive calls with the negation of domain constraints $D_1, \ldots, D_n$.
It $t$ has no polarity,
then we return the union of the domain constraints $D_1 \cup \ldots \cup D_n$.

@@

\begin{example}
Consider the (combined) signature $\Sigma = \Sigmalia \cup \Sigma_u$,
where $\ssorts{\Sigma_u} = \{ \sortint \}$ and 
$\sfuns{\Sigma_u} = \{ s\typ{ \sortint \rightarrow \sortint }, c\typ{ \sortint } \}$.
Consider the $\Sigma$-formula $\varphi$:
\begin{equation*}
\forall_{s} x\typ{\sortint}. ite( x \leq 0, s(x) \teq 0, s( x ) \teq x + s( x-1 ) ) \wedge s( c ) > 100
\end{equation*}
Here, $s$ is a function that returns the sum of positive numbers between $0$ and its argument $x$.
The formula $\varphi$ is well-defined with respect to functions $\sfundefs{\Sigma_u}$,
and states that the sum of all positive numbers between $0$ and $c$ is greater than $100$.
The signature $\extendsig{\Sigma}$ includes the sort $\fargsort{s}$,
and the uninterpreted function $\farg{s}$ of sort $\fargsort{s} \rightarrow \sortint$.
The result of $\conv( \varphi, \ptrue )$ is the pair $( \psi, \emptyset )$, where after simplification, $\psi$ is the $\extendsig{\Sigma}$-formula:
\begin{equation*}
\begin{split}
\forall y\typ{\fargsort{s}}. ite( & \farg{s}( y ) \leq 0,  \\
 & s(\farg{s}( y )) \teq 0, \\
 & s(\farg{s}( y )) \teq \farg{s}( y ) + s( \farg{s}( y )-1 ) ) \\
 & \wedge \exists z\typ{\fargsort{s}}. \farg{s}( z ) \teq \farg{s}( y )-1 )
\end{split}  
\wedge ( s( c ) > 100 \wedge \exists z\typ{\fargsort{s}}. \farg{s}( z ) \teq c )
\end{equation*}
The universal quantification in the resulting formula (the left disjunct) is over an uninterpreted sort $\fargsort{s}$, 
thus enabling finite model finding techniques~\cite{ReyEtAl-1-RR-13} to find a model for $\psi$ if a finite interpretation exists for $\fargsort{s}$.
Moreover, all occurrences of $y$ in this formula are beneath applications of the uninterpreted function $\farg{s}$,
implying that the formula is in the essentially uninterpreted fragment of~\cite{GeDeM-CAV-09}, for which there exists a complete instantiation procedure.
$\square$
\end{example}

It remains to show that our translation produces an equisatisfiable formulae.
Namely,
we claim that if $\varphi$ is well-defined with respect to functions $\sfundefs{\Sigma}$,
function definitions in $\sfundefs{\Sigma}$ are logically consistent \ajr{formalize},
then $\conv(\varphi)$ is equisatisfiable to $\varphi$.

\begin{lemma}
Let $\varphi$ be a $\Sigma$-formula that is well-defined with respect to functions $\sfundefs{\Sigma}$.
\ajr{todo}
\end{lemma}

\section{Evaluation}

In this section, we evaluate both 
the overall impact of the encoding introduced in the previous section, and
the performance of individual SMT techniques for benchmarks in the encoding.

For benchmarks, we gathered a set of 245 benchmarks from two sources, which we will refer to as {\bf isa} and {\bf leon}.
The first ({\bf isa}) consists of the 79 benchmarks from the Isaplanner benchmark suite~\cite{DBLP:conf/itp/JohanssonDB10} that did not contain higher-order functions.
The benchmarks have been used recently as challenge problems for a variety of inductive theorem provers.
These benchmarks heavily involve recursive functions, and are limited to the combined theory of uninterpreted functions and inductive datatypes.
The second ({\bf leon}) consists of 166 benchmarks from the Leon repository~\footnote{Available at \url{https://github.com/epfl-lara/leon}.},
which were constructed from verification conditions from simple Scala programs.
These benchmarks also heavily involve recursive functions, 
and cover a wide variety of theories, including bitvectors, arrays, linear and non-linear arithmetic, and inductive datatypes.
Each of the 245 benchmarks are of the form described in Formula (~\ref{eq:wdf}) from Section~\ref{sec:prelim},
namely they each are a conjoined list of function definitions followed by a (negated) conjecture.
A majority of these benchmarks are unsatisfiable, although a handful of the benchmarks in the {\bf leon} set are satisfiable.

For each of these 245 benchmarks, we considered (up to) three randomly selected mutated forms of its conjecture $\psi$.
In particular, we considered unique conjectures that are obtained as a result of swapping a subterm of $\psi$ at one position
with another of the same type at another position.
In total, we considered 213 mutated forms of conjectures from {\bf isa}, and 427 mutated forms of conjectures from {\bf leon}.
We will call these sets {\bf isa+m} and {\bf leon+m} respectively.

In total, our benchmark set consisted of 885 benchmarks (640 mutants and 245 originals).
We considered these 885 benchmarks both before and after the translation method $\conv$ as described in Section~\ref{sec:encoding}.
We will denote the latter encoding using the suffix {\bf w}, for instance {\bf isa+w} or {\bf isa+mw}.

For solvers, we considered the SMT solver \ziii~\cite{de-moura-bjoerner-2008}, 
which runs both heuristic methods for quantifier instantiation,
as well as methods for finding models for quantified formulas~\cite{GeDeM-CAV-09}.
We also considered three configurations of \cvc~\cite{barrett-et-al-2011} which we will refer to as {\bf cvc4}, {\bf cvc4+f} and {\bf cvc4+i}.
The default configuration {\bf cvc4} runs heuristic quantifier instantiation, but does not include techniques for finite model finding.
The configuration {\bf cvc4+f} runs a finite model finding procedure~\ref{ReyEtAl-1-RR-13}.
The last configuration {\bf cvc4+i} incorporates techniques for automating inductive reasoning in SMT~\ref{reynolds-kuncak-2015}.

The results are shown in Figure~\ref{fig:results}.

@@

{
\bibliographystyle{abbrv}
\bibliography{main}
}
